Lock-free Skiplist-based Priority Queues \\ Student Paper for Seminar in Algorithms 2013W \\ Technical University of Vienna

### 2014-03-11 paper/paper.tex
Sundell & Tsigas section

\begin{lstlisting}
struct slist_t {
  size_t max_level;
  node_t head[max_level];
};

struct node_t {
  key_t key;
  value_t value;
  size_t level;
  node_t next[level];
};

/* All operations are expected O(log n) time. */
void slist_insert(slist_t *l, key_t k, value_t v);
bool slist_delete(slist_t *l, key_t k, value_t &v);
\end{lstlisting}

\section{Fine-grained Locking Heaps}

@@

\subsection{Shavit \& Lotan}

@@

@@

\begin{lstlisting}
struct node_t {
  [...] /**< Standard node members as above. */
  atomic<bool> deleted; /**< Initially false. */
  time_t timestamp;
  lock_t locks[level + 1];
};
\end{lstlisting}

@@

@@

\subsection{Sundell \& Tsigas}

@@

In each test run, the examined priority queue was initially filled with $2^15$
elements. We then ran a tight loop of $50\%$ insertions and $50\%$ deletions
for a total of $10$ seconds, where all \lstinline|Insert|
operations within this context implicitly choose a key uniformly at random from
the range of all 32-bit integers. Each run was repeated for a total of $10$ times %TODO: Verify this.
and we report on the average throughput.

>>>>

\begin{figure}[h]
\begin{lstlisting}
struct slist_t {
  size_t max_level;
  node_t head[max_level];
};

struct node_t {
  key_t key;
  value_t value;
  size_t level;
  node_t *next[level];
};

/* All operations are expected O(log n) time. */
void slist_insert(slist_t *l, key_t k, value_t v);
bool slist_delete(slist_t *l, key_t k, value_t &v);
\end{lstlisting}
\caption{Basic SkipList structure and operations.}
\label{fig:basicsl}
\end{figure}

\section{Fine-grained Locking Heaps} \label{sec:hunt}

@@

\subsection{Shavit \& Lotan} \label{sec:shavit}

@@

% TODO: This counts for all skiplist impls.

@@

\begin{figure}[h]
\begin{lstlisting}
struct node_t {
  [...] /**< Standard node members as above. */
  atomic<bool> deleted; /**< Initially false. */
  time_t timestamp;
  lock_t locks[level + 1];
};
\end{lstlisting}
\caption{\citeauthor{shavit2000skiplist} structure.}
\label{fig:slsl}
\end{figure}

@@

% TODO: Linearizability points.

@@

\subsection{Sundell \& Tsigas} \label{sec:sundell}

\citeauthor{sundell2003fast} proposed the first lock-free concurrent priority queue in
\citeyear{sundell2003fast} \cite{sundell2003fast}. The data structure is linearizable
and is implemented using commonly available atomic primitives \ac{CAS}, \ac{TAS}, and \ac{FAA}.
In contrast to other structures covered in this paper, this priority queue is restricted to
contain items with pairwise distinct priorities.
A real-time version is also provided which will we will not discuss further (interested readers are
referred to \cite{sundell2003fast}).

\begin{figure}[h]
\begin{lstlisting}
struct node_t {
  [...] /**< Standard node members as above. */
  size_t valid_level;
  node_t *prev;
};
\end{lstlisting}
\caption{\citeauthor{sundell2003fast} structure.}
\label{fig:slsl}
\end{figure}

The structure of each node is basically identical to Figure \ref{fig:basicsl}. However, \citeauthor{sundell2003fast}
exploit the fact that the two least significant bits of pointers on 32- and 64-bit systems
are unused and reuse pointers as deletion marks. A set least significant bit signifies on a pointer signifies
that the current node is about to be deleted. We will use the notation 
\lstinline|CAS(node.level[i], (ptr, false), (ptr, true))| to signify \ac{CAS} operations on the
combined pointer \& flag fields.
Reuse of \lstinline|node.level[i]| pointers
prevents situations in which a new node is inserted while its predecessor is being removed,
effectively deleting both nodes from the list. Likewise, the reuse of the \lstinline|node.value|
pointer ensures that updates of pointer values (which occur when a node with the inserted priority already exists)
handle concurrent node removals correctly.

As in the \citeauthor{shavit2000skiplist} priority queue, insertions proceed bottom-up while
deletions proceed top-down --- on the one hand, the choice of opposite directions reduces collisions
between concurrent insert and delete operations, while on the other hand removing nodes from top levels first
allows many other concurrent operations to simply skip these nodes, further improving performance.
\lstinline|node.valid_level| is updated during inserts to always equal the highest level of the SkipList
at which pointers in this node have already been set (as opposed to \lstinline|node.level|, which equals
the final level of the node).

A helping mechanism is employed whenever a node is encountered that has its deletion bit set, which attempts
to set the deletion bits on all next pointers and then removes the node from the current level. The
\lstinline|node.prev| pointer is used as a shortcut to the previous node, avoiding a complete retraversal
of the list.

This implementation uses the lock-free memory management by invented by \citeauthor{valois1996lock}
\cite{valois1995lock,valois1996lock} and corrected by \citeauthor{michael1995correction} 
\cite{michael1995correction}. It was chosen in particular because this scheme can guarantee validity
of \lstinline|prev| as well as all \lstinline|next| pointers. Additionally, it does not require a separate
garbage collector thread.

A rigorous linearizability proof is provided in the original paper \cite{sundell2003fast} which shows
linearizability points for all possible outcomes of both \lstinline|Insert| and \lstinline|DeleteMin|
operations.

% TODO: Further details?

Benchmarks performed by \citeauthor{sundell2003fast} show their queue performing noticeably better than both locking 
queues from Sections \ref{sec:shavit} and \ref{sec:hunt}, and slightly better than a priority queue
consisting of a SkipList protected by a single global lock.

@@

In each test run, the examined priority queue was initially filled with $2^{15}$
elements. We then ran a tight loop of $50\%$ insertions and $50\%$ deletions
for a total of $10$ seconds, where all \lstinline|Insert|
operations within this context implicitly choose a key uniformly at random from
the range of all 32-bit integers. Each run was repeated for a total of $10$ times %TODO: Verify this.
and we report on the average throughput.
